Flow-Based Cost Allocation in Linear Optimized Power Systems are non-unique but localizable

### 2020-07-21 paper/flow_based.tex
paper rename newcommands, finish brownfield section

% general symbols
\newcommand{\generation}{g_{s,t}}
\newcommand{\generationpotential}{\bar{g}_{s,t}}
\newcommand{\generationshare}[1][n]{\omega_{#1,s,t}}
\newcommand{\nodalgeneration}[1][n]{g_{#1,t}}
\newcommand{\capacityGeneration}{G_{s}}
\newcommand{\capacityGenerationUpper}{\bar{G}_{s}}
\newcommand{\capacityGenerationLower}{\ubar{G}_{s}}
\newcommand{\capacityFlow}{F_{\ell}}
\newcommand{\capacityFlowUpper}{\bar{F}_{\ell}}
\newcommand{\capacityFlowLower}{\ubar{F}_{\ell}}
\newcommand{\capexGeneration}{c_{s}}
\newcommand{\capexFlow}{c_{\ell}}
\newcommand{\opexGeneration}{o_{s}}
\newcommand{\demand}[1][n]{d_{#1,t}}
\newcommand{\nodaldemand}[1][n]{d_{#1,t}}
\newcommand{\demandshare}[1][n]{\omega_{#1,a,t}}
\newcommand{\utility}{U_{n,a,t}}
\newcommand{\incidence}[1][n]{K_{#1,\ell}}
\newcommand{\incidenceGenerators}[1][n]{K_{#1,s}}
\newcommand{\ptdf}[1][n]{H_{\ell,#1}}
\newcommand{\ptdfEqual}[1][n]{\ptdf[#1]^\circ}
\newcommand{\slackflow}{k_{\ell}}
\newcommand{\slack}[1][n]{k_{#1}}
\newcommand{\slackk}[1][n]{k^*_{#1}}
\newcommand{\Slack}{k_{m,n}}
\newcommand{\Slackk}{k^*_{m,n}}
\newcommand{\mulowergeneration}{\ubar{\mu}_{s,t}}
\newcommand{\muuppergeneration}{\bar{\mu}_{s,t}}
\newcommand{\muuppergenerationnom}{\bar{\mu}^\text{nom}_{s}}
\newcommand{\mulowergenerationnom}{\ubar{\mu}^\text{nom}_{s}}
\newcommand{\mulowerflow}{\ubar{\mu}_{\ell,t}}
\newcommand{\muupperflow}{\bar{\mu}_{\ell,t}}
\newcommand{\muupperflownom}{\bar{\mu}^\text{nom}_{\ell}}
\newcommand{\mulowerflownom}{\ubar{\mu}^\text{nom}_{\ell}}
\newcommand{\lmp}[1][n]{\lambda_{#1,t}}
\newcommand{\flow}{f_{\ell,t}}
\newcommand{\cycle}{C_{\ell,c}}
\newcommand{\impedance}{x_\ell}
\newcommand{\cycleprice}{\lambda_{c,t}}
\newcommand{\injection}{p_{n,t}}
\newcommand{\netconsumption}[1][n]{p^{-}_{#1,t}}
\newcommand{\netproduction}[1][n]{p^{+}_{#1,t}}
\newcommand{\selfconsumption}[1][n]{p^{\circ}_{#1,t}}

@@

% allocation quantities
\newcommand{\allocatePeer}[1][s \rightarrow n]{A_{#1,t}}
\newcommand{\allocateFlow}[1][n]{A_{\ell,#1,t}}
\newcommand{\allocateTransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocateCapexGeneration}[1][n]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocateCapexFlow}[1][n]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocateOpex}[1][n]{\mathcal{O}_{#1,t}}
\newcommand{\allocateEmissionCost}[1][n]{\mathcal{E}_{#1,t}}

\newcommand{\emission}{e_{s}}
\newcommand{\emissionPrice}{\mu_{\text{CO2}}}
\newcommand{\megawatthour}{MWh$_\text{el}$}
\newcommand{\totalcost}{\mathcal{TC}}
\newcommand{\totalOpexGeneration}{\mathcal{O}}
\newcommand{\totalOpexFlow}{\mathcal{O}^F}
\newcommand{\totalCapexGeneration}{\mathcal{C}^G}
\newcommand{\totalCapexFlow}{\mathcal{C}^F}
\newcommand{\totalEmissionCost}{\mathcal{E}}
\newcommand{\totalRest}{\mathcal{R}}

@@

% In the following, we show how Flow Allocation (FA) can be used for decomposing Locational Marginal Prices (LMP) in a linear optimized power system. We begin by revising the fundamentals of the Power Transfer Distribition Factors (PTDF) and the role of the slack, which plays a key role for FA. This will lead us to a generalized form of P2P allocations and transmission usage allocations based on the choice of slack. In a second step, we breakdown the full Lagrangian of a cost-optimization for a network with transmission and generation capacity expansion. By taking into account sensitivities of the flow with respect to changes in the nodal generation, we are able to derive a cost allocation which directly links to FA.

@@

\begin{table}[h]
    \centering
    \begin{tabular}{ll}
        $\lmp$ & Locational Market Price at bus $n$ and  \\ & time step $t$ in \euro/MW \vpad \\
        $\demand$ & Electric demand per bus $n$, demand type $a$, \\ & time step $t$ in MW  \vpad \\
        $\generation$ & Electric generation of generator $s$, \\ & time step $t$  in MW \vpad \\
        $\flow$ & Active power flow on line $\ell$, \\ & time step $t$ in MW   \vpad \\
        $\opexGeneration$ & Operational price in \euro/MW \vpad \\
        $\capexGeneration$ & Capital Price in \euro/MW \vpad \\
        $\capexFlow$  & Capital Price in \euro/MW for transmission \\ & capacity on line $\ell$  \vpad \\
        $\capacityGeneration$ & Generation capacity in MW \vpad \\
        $\capacityFlow$ & Transmission capacity in MW \vpad \\
        $\incidence$ & Incidence matrix \vpad 
    \end{tabular}
\end{table}

@@

In long-term operation and investment planning models, the total costs $\totalcost$ of a power system include all operational expenditures (OPEX) $\totalOpexGeneration$ and capital expenditures (CAPEX) $\totalCapexGeneration$ and $\totalCapexFlow$ for production  and transmission respectively, thus
\begin{align}
\totalcost &= \totalOpexGeneration + \totalEmissionCost +  \totalCapexGeneration +  \totalCapexFlow + ...
\label{eq:total_cost_terms}
\end{align}
In a network design with minimal $\totalcost$, the Locational Market Price (LMP) describes the marginal price for an incremental increase of electricity demand $\demand$ of any consumer $a$ at bus $n$. It is given by the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$
\begin{align}
\lmp = \pdv{\,\totalcost}{\,\demand}
\label{eq:total_cost_derivative}
\end{align}
% 
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{price_decomposition.png}
\caption{Schematic decomposition of the Locational Market Price $\lmp$. In power system model with optimal long-term operation and planning, the total system costs $\totalcost$ split into different cost terms, \ie OPEX and CAPEX for production and transmission and possibly other expenditures. }
\label{fig:price_decomposition}
\end{figure}
%

This leads to a nodal pricing where over the span of optimized timesteps $t$, the system costs are partially or totally payed back by the consumers 
\begin{align}
\totalcost - \totalRest &=  \sum_{n,a,t} \lmp \, \demand
\label{eq:total_cost_sum}
\end{align}
depending on the costs $\totalRest$ which are independent of the nodal demand  
\begin{align}
 \pdv{\totalRest}{\demand} = 0
\end{align}
% 
In most cases, where $\totalRest$ plays a minor role, the LMP splits into different cost terms assigned to OPEX and CAPEX for production and transmission and eventual other expenditures. We schematically show this behaviour in \cref{fig:price_decomposition}. 
Extensive investigations of the LMP as done in \cite{schweppe_spot_1988} already showed this connection however leave the question open how the costs are allocated among the components. In the following we will show how the above presented cost decomposition is extended to full peer-to-peer (P2P) cost allocation including all network components.

@@

From inserting \cref{eq:total_cost_terms} into \cref{eq:total_cost_derivative}, we are able to disaggregate the cost terms $\totalOpexGeneration$, $\totalCapexGeneration$ and $\totalCapexFlow$ into constributions of demand $\demand$ at bus~$n$ and time~$t$. For the CAPEX allocation we will make use of the shadowprices resulting from the cost mimization.

@@

The total OPEX is given by the operational price $\opexGeneration$ times the generation $\generation$ of all generators $s$, 
\begin{align}
 \totalOpexGeneration& = \sum_{s,t} \opexGeneration\,  \generation 
\label{eq:opex_generation}
 \end{align}
The allocated OPEX $\allocateOpex$ which naturally sum up to $  \totalOpexGeneration = \sum_{n,t} \allocateOpex$, are given by 
\begin{align}
\allocateOpex  = \pdv{\totalOpexGeneration}{\demand} \, \demand & =  \sum_{s,t}\opexGeneration \, \pdv{\generation}{\demand} \, \demand
\end{align}
% 
The derivative on the right hand side time the nodal demand can be interpreted as power produced by generator $s$ and consumed at node $n$. Without further explaining how the derivative is calculated, we define this power allocation as 
\begin{align}
 \allocatePeer = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
% 
which by feeding back into \cref{eq:opex_generation} has to fulfill
\begin{align}
\generation = \sum_n \allocatePeer
\label{eq:allocate_peer_constraint}
\end{align}
% 
% 
The direct payment of node $n$ to generator $s$ is then given by  
% 
\begin{align}
 \allocateOpex[n \rightarrow s] &= 
\opexGeneration \,  \allocatePeer
\label{eq:allocate_opexGeneration_detailed}
\end{align}
Accordingly, consumers at bus $n$ which retrieve power from a generators $s$, directly  compensate its operational costs. The cost allocation behaves like P2P tradings between producers and consumers for fixed production prices. In this way,  generator $s$ retrieves the exact amount of money from consumers that it spends on its operation, thus 
\begin{align}
\sum_{n} \allocateOpex[n \rightarrow s] = \opexGeneration \, \generation
\label{eq:no_profit_opex}
\end{align}

@@

Given a fix price for emissions in CO$_2$ equivalents $\emissionPrice$, the cost term for emission adds up to 
\begin{align}
 \totalEmissionCost = \emissionPrice \, \sum_s  \emission \, \generation
\end{align}
where $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour for generator $s$.
Similar to the allocated OPEX, the emission cost for consumers at bus $n$ at time $t$ is allocated through
\begin{align}
 \allocateEmissionCost = \pdv{\totalEmissionCost}{\demand} \demand = \sum_s \emissionPrice \, \emission \, \allocatePeer  
\end{align}
with its peer-to-peer payments 
\begin{align}
 \allocateEmissionCost[n \rightarrow s] = \emissionPrice \, \emission \, \allocatePeer
\end{align}

@@

The optimization of the generation capacity $\capacityGeneration$ adds the cost term 
\begin{align}
 \totalCapexGeneration& = \sum_{s} \capexGeneration \capacityGeneration 
\end{align}
to $\totalcost$. It combines all capacities times their capital cost $\capexGeneration$. The nominal capacity constraints the generation $\generation$ in the form of 
\begin{align}
\generation - \generationpotential \capacityGeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{n,s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}\\ 
- \generation &\le 0 \resultsin{\mulowergeneration} \Forall{n,s,t} 
\label[constraint]{eq:lower_generation_capacity_constraint}
\end{align}
where $\generationpotential \in \left[ 0,1\right]$ is the capacity factor for renewable generators. Using the result of \cite{brown_decreasing_2020} we show in that the capital cost  for generator $s$ translates is payed back by the generation times the shadow price for the upper capacity constraint, 
\begin{align}
 \capexGeneration \capacityGeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
With our definition in \cref{eq:allocate_peer} the allocated CAPEX can then be defined as 
\begin{align}
\allocateCapexGeneration = \pdv{\totalCapexGeneration}{\demand} \, \demand & = \sum_{s} \muuppergeneration \, \allocatePeer
\end{align}
with the peer-to-peer allocation 
\begin{align}
 \allocateCapexGeneration[n \rightarrow s] = \muuppergeneration \, \allocatePeer
 \label{eq:allocate_capexGeneration_detailed}
\end{align}

So, how does this allocation behave? According to the polluter pays principle, it differentiates between consumers who are `responsible` for investments and those who are not. If $\muuppergeneration$ (in literature often denoted as the Quality of Supply) is bigger than zero, the upper Capacity \cref{eq:upper_generation_capacity_constraint} is binding. Thus it is these times steps which push investments in $\capacityGeneration$. If $\muuppergeneration = 0$, the generation $\generation$ is not bound and investments are not necessary. 
When summing over all CAPEX payments to generator $s$ in \cref{eq:allocate_capexGeneration_detailed} and using \cref{eq:allocate_peer_constraint,eq:no_profit_capex_generation}, we see that each generator retrieves exactly the cost that were spent to build the capacity $\capacityGeneration$.

@@

The CAPEX for the transmission system is given by 
\begin{align}
 \totalCapexFlow & = \sum_\ell \capexFlow \capacityFlow
 \label{eq:capex_flow}
\end{align}
The transmission capacity variable $\capacityFlow$ limits the flow $\flow$ in both directions, such that 
\begin{align}
\flow - \capacityFlow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityFlow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
Again, we use the result of \cite{brown_decreasing_2020} which derives that the investment in line $\ell$ is payed back by the shadowprices times the flow 
\begin{align}
\capexFlow \capacityFlow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
We insert this into \cref{eq:capex_flow} and define the individual cost terms  as 
\begin{align}
\allocateCapexFlow = \pdv{\totalCapexFlow}{\demand}\demand & =  \sum_{\ell,t} \left( \muupperflow - \mulowerflow \right) \pdv{\flow}{\demand} \, \demand
\end{align}
The marginal change of flow with repect to the nodal demand is again non-unique and requires assumption on where the power is resulting from.

\begin{align}
 \pdv{\flow}{\demand}\, \demand = \allocateFlow
\end{align}
fulfilling 
\begin{align}
\sum_{n} \allocateFlow = \flow 
\label{eq:allocate_transaction_constraint}
\end{align}
% 
The peer-to-peer cost allocation for the transmission CAPEX is then defined as 
\begin{align}
 \allocateCapexFlow[n \rightarrow \ell] &=  
\left( \muupperflow - \mulowerflow\right)  \allocateFlow
\label{eq:allocate_capexFlow_detailed}
\end{align}
% 
Again we can interpret the shadowprices $\muupperflow$ and $\mulowerflow$ as a measure for necessity of transmission investments at $\ell$ at time $t$. Consumers which retrieve power flowing on congested lines, yielding a bound \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint}, pay compensations for the resulting investments. Again the sum of all CAPEX payments to line $\ell$ equals the total CAPEX spent. This is seen when summing \cref{eq:allocate_capexFlow_detailed} over all buses and time steps and using \cref{eq:allocate_transaction_constraint,eq:no_profit_capex_flow}

@@

The P2P cost allocation suits for any type of topology and network setup. But so far, the question of how $\allocatePeer$ and $\allocateFlow$ are defined was left open. We recap that both rely on derivatives of $\generation$ and $\flow$  with respect to the nodal demand $\demand$ fulfilling \cref{eq:allocate_peer_constraint,eq:allocate_transaction_constraint}. \\
% 
In fact, it turns out that ones the peer-to-peer allocations $\allocatePeer$ are defined, the power flow allocations $\allocateFlow$ are determined by the linear power flow equations.
In order to show this, let $\ptdf$ denote Power Transfer Distribition Factors (PTDF) giving the changes in the flow on line $\ell$ for one unit (typically one MW) of net power production at bus $n$.  Further let $\incidenceGenerators$ be 1 if generator $s$ is attached to bus $n$ and zero otherwise. The linear power flow equals the 
\begin{align}
 \flow  = \sum_m \ptdf[m] \left( \nodalgeneration[m] - \demand[m] \right)  
\end{align}
where the nodal generation $\nodalgeneration[m] = \sum_s \incidenceGenerators[m] \, \generation $ combines the production of all generators attached to bus $m$. Taking the derivative with respect to the demand, leads to 
\begin{align}
 \allocateFlow = \pdv{\flow}{\demand} \demand = \sum_m \ptdf[m] \left( \allocatePeer[m \rightarrow n]  - \delta_{n,m} \demand \right) 
 \label{eq:allocate_peer_to_allocate_flow}
\end{align}
where $\allocatePeer[m \rightarrow n] = \sum_s \incidenceGenerators \, \allocatePeer$ combines all power allocations of the source $m$. \Cref{eq:allocate_peer_to_allocate_flow} shows that the peer-to-peer allocation $\allocatePeer$ directly determines the flow allocation on line $\ell$.

>>>>

% general symbols
\newcommand{\generation}{g_{s,t}}
\newcommand{\generationpotential}{\bar{g}_{s,t}}
\newcommand{\generationshare}[1][n]{\omega_{#1,s,t}}
\newcommand{\nodalgeneration}[1][n]{g_{#1,t}}
\newcommand{\capacitygeneration}{G_{s}}
\newcommand{\capacitygenerationupper}{\bar{G}_{s}}
\newcommand{\capacitygenerationlower}{\ubar{G}_{s}}
\newcommand{\capacityflow}{F_{\ell}}
\newcommand{\capacityflowUpper}{\bar{F}_{\ell}}
\newcommand{\capacityflowLower}{\ubar{F}_{\ell}}
\newcommand{\capitalpricegeneration}{c_{s}}
\newcommand{\capitalpriceflow}{c_{\ell}}
\newcommand{\operationalpricegeneration}{o_{s}}
\newcommand{\demand}[1][n]{d_{#1,t}}
\newcommand{\nodaldemand}[1][n]{d_{#1,t}}
\newcommand{\demandshare}[1][n]{\omega_{#1,a,t}}
\newcommand{\utility}{U_{n,a,t}}
\newcommand{\incidence}[1][n]{K_{#1,\ell}}
\newcommand{\incidencegenerator}[1][n]{K_{#1,s}}
\newcommand{\ptdf}[1][n]{H_{\ell,#1}}

\newcommand{\mulowergeneration}{\ubar{\mu}_{s,t}}
\newcommand{\muuppergeneration}{\bar{\mu}_{s,t}}
\newcommand{\muuppergenerationnom}{\bar{\mu}^\text{nom}_{s}}
\newcommand{\mulowergenerationnom}{\ubar{\mu}^\text{nom}_{s}}
\newcommand{\mulowerflow}{\ubar{\mu}_{\ell,t}}
\newcommand{\muupperflow}{\bar{\mu}_{\ell,t}}
\newcommand{\muupperflownom}{\bar{\mu}^\text{nom}_{\ell}}
\newcommand{\mulowerflownom}{\ubar{\mu}^\text{nom}_{\ell}}
\newcommand{\lmp}[1][n]{\lambda_{#1,t}}
\newcommand{\flow}{f_{\ell,t}}
\newcommand{\cycle}{C_{\ell,c}}
\newcommand{\impedance}{x_\ell}
\newcommand{\cycleprice}{\lambda_{c,t}}
\newcommand{\injection}{p_{n,t}}
\newcommand{\netconsumption}[1][n]{p^{-}_{#1,t}}
\newcommand{\netproduction}[1][n]{p^{+}_{#1,t}}
\newcommand{\selfconsumption}[1][n]{p^{\circ}_{#1,t}}

@@

% allocation quantities
\newcommand{\allocatepeer}[1][s \rightarrow n]{A_{#1,t}}
\newcommand{\allocateflow}[1][n]{A_{\ell,#1,t}}
\newcommand{\allocatetransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocatecapexgeneration}[1][n]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocatecapexflow}[1][n]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocateopex}[1][n]{\mathcal{O}_{#1,t}}
\newcommand{\allocateemissioncost}[1][n]{\mathcal{E}_{#1,t}}

\newcommand{\emission}{e_{s}}
\newcommand{\emissionprice}{\mu_{\text{CO2}}}
\newcommand{\megawatthour}{MWh$_\text{el}$}
\newcommand{\totalcost}{\mathcal{TC}}
\newcommand{\opexgeneration}{\mathcal{O}}
\newcommand{\opexflow}{\mathcal{O}^F}
\newcommand{\capexgeneration}{\mathcal{C}^G}
\newcommand{\capexflow}{\mathcal{C}^F}
\newcommand{\emissioncost}{\mathcal{E}}
\newcommand{\remainingcost}{\mathcal{R}}

@@

@@

\begin{table}[h]
    \centering
    \begin{tabular}{ll}
        $\lmp$ & Locational Market Price at bus $n$ and  \\ & time step $t$ in \euro/MW \vpad \\
        $\demand$ & Electric demand per bus $n$, demand type $a$, \\ & time step $t$ in MW  \vpad \\
        $\generation$ & Electric generation of generator $s$, \\ & time step $t$  in MW \vpad \\
        $\flow$ & Active power flow on line $\ell$, \\ & time step $t$ in MW   \vpad \\
        $\operationalpricegeneration$ & Operational price in \euro/MW \vpad \\
        $\capitalpricegeneration$ & Capital Price in \euro/MW \vpad \\
        $\capitalpriceflow$  & Capital Price in \euro/MW for transmission \\ & capacity on line $\ell$  \vpad \\
        $\capacitygeneration$ & Generation capacity in MW \vpad \\
        $\capacityflow$ & Transmission capacity in MW \vpad \\
        $\incidence$ & Incidence matrix \vpad 
    \end{tabular}
\end{table}

@@

In long-term operation and investment planning models, the total costs $\totalcost$ of a power system include all operational expenditures (OPEX) $\opexgeneration$ and capital expenditures (CAPEX) $\capexgeneration$ and $\capexflow$ for production  and transmission respectively, thus
\begin{align}
\totalcost &= \opexgeneration + \emissioncost +  \capexgeneration +  \capexflow + ...
\label{eq:total_cost_terms}
\end{align}
In a network design with minimal $\totalcost$, the Locational Market Price (LMP) describes the marginal price for an incremental increase of electricity demand $\demand$ of any consumer $a$ at bus $n$. It is given by the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$
\begin{align}
\lmp = \pdv{\,\totalcost}{\,\demand}
\label{eq:total_cost_derivative}
\end{align}
% 
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{price_decomposition.png}
\caption{Schematic decomposition of the Locational Market Price $\lmp$. In power system model with optimal long-term operation and planning, the total system costs $\totalcost$ split into different cost terms, \ie OPEX and CAPEX for production and transmission and possibly other expenditures. }
\label{fig:price_decomposition}
\end{figure}
%

This leads to a nodal pricing where over the span of optimized timesteps $t$, the system costs are partially or totally payed back by the consumers 
\begin{align}
\totalcost - \remainingcost &=  \sum_{n,a,t} \lmp \, \demand
\label{eq:total_cost_sum}
\end{align}
depending on the costs $\remainingcost$ which are independent of the nodal demand  
\begin{align}
 \pdv{\remainingcost}{\demand} = 0
\end{align}
% 
In most cases, where $\remainingcost$ plays a minor role, the LMP splits into different cost terms assigned to OPEX and CAPEX for production and transmission and eventual other expenditures. We schematically show this behaviour in \cref{fig:price_decomposition}. 
Extensive investigations of the LMP as done in \cite{schweppe_spot_1988} already showed this connection however leave the question open how the costs are allocated among the components. In the following we will show how the above presented cost decomposition is extended to full peer-to-peer (P2P) cost allocation including all network components.

@@

From inserting \cref{eq:total_cost_terms} into \cref{eq:total_cost_derivative}, we are able to disaggregate the cost terms $\opexgeneration$, $\capexgeneration$ and $\capexflow$ into constributions of demand $\demand$ at bus~$n$ and time~$t$. For the CAPEX allocation we will make use of the shadowprices resulting from the cost mimization.

@@

The total OPEX is given by the operational price $\operationalpricegeneration$ times the generation $\generation$ of all generators $s$, 
\begin{align}
 \opexgeneration& = \sum_{s,t} \operationalpricegeneration\,  \generation 
\label{eq:opex_generation}
 \end{align}
The allocated OPEX $\allocateopex$ which naturally sum up to $  \opexgeneration = \sum_{n,t} \allocateopex$, are given by 
\begin{align}
\allocateopex  = \pdv{\opexgeneration}{\demand} \, \demand & =  \sum_{s,t}\operationalpricegeneration \, \pdv{\generation}{\demand} \, \demand
\end{align}
% 
The derivative on the right hand side time the nodal demand can be interpreted as power produced by generator $s$ and consumed at node $n$. Without further explaining how the derivative is calculated, we define this power allocation as 
\begin{align}
 \allocatepeer = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
% 
which by feeding back into \cref{eq:opex_generation} has to fulfill
\begin{align}
\generation = \sum_n \allocatepeer
\label{eq:allocate_peer_constraint}
\end{align}
% 
% 
The direct payment of node $n$ to generator $s$ is then given by  
% 
\begin{align}
 \allocateopex[n \rightarrow s] &= 
\operationalpricegeneration \,  \allocatepeer
\label{eq:allocate_opexGeneration_detailed}
\end{align}
Accordingly, consumers at bus $n$ which retrieve power from a generators $s$, directly  compensate its operational costs. The cost allocation behaves like P2P tradings between producers and consumers for fixed production prices. In this way,  generator $s$ retrieves the exact amount of money from consumers that it spends on its operation, thus 
\begin{align}
\sum_{n} \allocateopex[n \rightarrow s] = \operationalpricegeneration \, \generation
\label{eq:no_profit_opex}
\end{align}

@@

Given a fix price for emissions in CO$_2$ equivalents $\emissionprice$, the cost term for emission adds up to 
\begin{align}
 \emissioncost = \emissionprice \, \sum_s  \emission \, \generation
\end{align}
where $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour for generator $s$.
Similar to the allocated OPEX, the emission cost for consumers at bus $n$ at time $t$ is allocated through
\begin{align}
 \allocateemissioncost = \pdv{\emissioncost}{\demand} \demand = \sum_s \emissionprice \, \emission \, \allocatepeer  
\end{align}
with its peer-to-peer payments 
\begin{align}
 \allocateemissioncost[n \rightarrow s] = \emissionprice \, \emission \, \allocatepeer
\end{align}

@@

The optimization of the generation capacity $\capacitygeneration$ adds the cost term 
\begin{align}
 \capexgeneration& = \sum_{s} \capitalpricegeneration \capacitygeneration 
\end{align}
to $\totalcost$. It combines all capacities times their capital cost $\capitalpricegeneration$. The nominal capacity constraints the generation $\generation$ in the form of 
\begin{align}
\generation - \generationpotential \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{n,s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}\\ 
- \generation &\le 0 \resultsin{\mulowergeneration} \Forall{n,s,t} 
\label[constraint]{eq:lower_generation_capacity_constraint}
\end{align}
where $\generationpotential \in \left[ 0,1\right]$ is the capacity factor for renewable generators. Using the result of \cite{brown_decreasing_2020} we show in that the capital cost  for generator $s$ translates is payed back by the generation times the shadow price for the upper capacity constraint, 
\begin{align}
 \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
With our definition in \cref{eq:allocate_peer} the allocated CAPEX can then be defined as 
\begin{align}
\allocatecapexgeneration = \pdv{\capexgeneration}{\demand} \, \demand & = \sum_{s} \muuppergeneration \, \allocatepeer
\end{align}
with the peer-to-peer allocation 
\begin{align}
 \allocatecapexgeneration[n \rightarrow s] = \muuppergeneration \, \allocatepeer
 \label{eq:allocate_capexGeneration_detailed}
\end{align}

So, how does this allocation behave? According to the polluter pays principle, it differentiates between consumers who are `responsible` for investments and those who are not. If $\muuppergeneration$ (in literature often denoted as the Quality of Supply) is bigger than zero, the upper Capacity \cref{eq:upper_generation_capacity_constraint} is binding. Thus it is these times steps which push investments in $\capacitygeneration$. If $\muuppergeneration = 0$, the generation $\generation$ is not bound and investments are not necessary. 
When summing over all CAPEX payments to generator $s$ in \cref{eq:allocate_capexGeneration_detailed} and using \cref{eq:allocate_peer_constraint,eq:no_profit_capex_generation}, we see that each generator retrieves exactly the cost that were spent to build the capacity $\capacitygeneration$.

@@

The CAPEX for the transmission system is given by 
\begin{align}
 \capexflow & = \sum_\ell \capitalpriceflow \capacityflow
 \label{eq:capex_flow}
\end{align}
The transmission capacity variable $\capacityflow$ limits the flow $\flow$ in both directions, such that 
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
Again, we use the result of \cite{brown_decreasing_2020} which derives that the investment in line $\ell$ is payed back by the shadowprices times the flow 
\begin{align}
\capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
We insert this into \cref{eq:capex_flow} and define the individual cost terms  as 
\begin{align}
\allocatecapexflow = \pdv{\capexflow}{\demand}\demand & =  \sum_{\ell,t} \left( \muupperflow - \mulowerflow \right) \pdv{\flow}{\demand} \, \demand
\end{align}
The marginal change of flow with repect to the nodal demand is again non-unique and requires assumption on where the power is resulting from.

\begin{align}
 \pdv{\flow}{\demand}\, \demand = \allocateflow
\end{align}
fulfilling 
\begin{align}
\sum_{n} \allocateflow = \flow 
\label{eq:allocate_transaction_constraint}
\end{align}
% 
The peer-to-peer cost allocation for the transmission CAPEX is then defined as 
\begin{align}
 \allocatecapexflow[n \rightarrow \ell] &=  
\left( \muupperflow - \mulowerflow\right)  \allocateflow
\label{eq:allocate_capexFlow_detailed}
\end{align}
% 
Again we can interpret the shadowprices $\muupperflow$ and $\mulowerflow$ as a measure for necessity of transmission investments at $\ell$ at time $t$. Consumers which retrieve power flowing on congested lines, yielding a bound \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint}, pay compensations for the resulting investments. Again the sum of all CAPEX payments to line $\ell$ equals the total CAPEX spent. This is seen when summing \cref{eq:allocate_capexFlow_detailed} over all buses and time steps and using \cref{eq:allocate_transaction_constraint,eq:no_profit_capex_flow}

@@

The P2P cost allocation suits for any type of topology and network setup. But so far, the question of how $\allocatepeer$ and $\allocateflow$ are defined was left open. We recap that both rely on derivatives of $\generation$ and $\flow$  with respect to the nodal demand $\demand$ fulfilling \cref{eq:allocate_peer_constraint,eq:allocate_transaction_constraint}. \\
% 
In fact, it turns out that ones the peer-to-peer allocations $\allocatepeer$ are defined, the power flow allocations $\allocateflow$ are determined by the linear power flow equations.
In order to show this, let $\ptdf$ denote Power Transfer Distribition Factors (PTDF) giving the changes in the flow on line $\ell$ for one unit (typically one MW) of net power production at bus $n$.  Further let $\incidencegenerator$ be 1 if generator $s$ is attached to bus $n$ and zero otherwise. The linear power flow equals the 
\begin{align}
 \flow  = \sum_m \ptdf[m] \left( \nodalgeneration[m] - \demand[m] \right)  
\end{align}
where the nodal generation $\nodalgeneration[m] = \sum_s \incidencegenerator[m] \, \generation $ combines the production of all generators attached to bus $m$. Taking the derivative with respect to the demand, leads to 
\begin{align}
 \allocateflow = \pdv{\flow}{\demand} \demand = \sum_m \ptdf[m] \left( \allocatepeer[m \rightarrow n]  - \delta_{n,m} \demand \right) 
 \label{eq:allocate_peer_to_allocate_flow}
\end{align}
where $\allocatepeer[m \rightarrow n] = \sum_s \incidencegenerator \, \allocatepeer$ combines all power allocations of the source $m$. \Cref{eq:allocate_peer_to_allocate_flow} shows that the peer-to-peer allocation $\allocatepeer$ directly determines the flow allocation on line $\ell$.
